A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-01-03 analysis/paper/paper.Rmd
Re-running analyses and writing

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @solheim2020; @åstveit2018]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves. The isobases have a direction of 327$^{\\circ}$ (Romundset el al. 2018), B) Displacement curves. Note the increasing steepness of the curves towards the north-east."), fig.height = 13}
bboxsites <- st_bbox(sites_sa)
bboxsites[1] <- bboxsites[1] - 15000
bboxsites[3] <- bboxsites[3] + 15000
bboxsites[2] <- bboxsites[2] - 5000
bboxsites[4] <- bboxsites[4] + 5000
bboxsitespoly <- st_as_sf(st_as_sfc(bboxsites))

@@

anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
overview_map <- ggplot() +
  geom_sf(data = bmap_reproj, fill = "#fbecdc", colour = "black") +
  geom_sf(data = isobases, aes(colour = name)) +
  geom_sf(data = st_centroid(sites_sa), fill = "red",
          size = 2.25, shape = 21,
          colour = "black") +
  ggsn::scalebar(data = sites_sa, dist = 20, dist_unit = "km",
                 transform = FALSE, st.size = 4, height = 0.02,
                 border.size = 0.1, st.dist = 0.03,
                 anchor = c(x = anc[2] - 15500, y = anc[1]) + 8000) +
  coord_sf(xlim = c(bboxsites[1], bboxsites[3]),
           ylim = c(bboxsites[2], bboxsites[4]),
           expand = FALSE) +
  scale_colour_manual(values = c("Arendal" = "black",
                                   "Larvik" = "darkgreen",
                                   "Tvedestrand" = "blue",
                                   "Horten" = "darkorange")) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     panel.background = element_rect(fill = "#dbe3f3", 
                                                    colour = "black"),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

@@

To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done for each site by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. These values were then subtracted from the dsiplacement curve located to the south-west of the site. The result of this process is shown for an example site in Figure \@ref(fig:example-1).

@@

# Plot the drawn elevation
elevsamp <- ggplot(sitecurve, aes(x = years)) +
  geom_line(aes(y =  upperelev), colour = "red") +
  geom_line(aes(y = lowerelev), colour = "red") +
  ylab("Meters above present sea-level") +
  xlab("cal BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal") +
  geom_segment(aes(x = sdate, xend = sdate, y = uppers,
                                           yend = -Inf), col = "black",
                size = 0.25, linetype = "solid") +
  geom_segment(aes(x = -Inf, xend = sdate, y = uppers,
                   yend = uppers), col = "black", size = 0.25) +
  geom_segment(aes(x = -Inf, xend = sdate, y = lowers,
                   yend = lowers), col = "black", size = 0.25) +
  geom_segment(aes(x = -Inf, xend = sdate, y = lowers,
                   yend = lowers), col = "black", size = 0.25) +
  geom_segment(aes(x = -Inf, xend = sdate, y = sealevel,
                   yend = sealevel), col = "red", size = 0.25,
               linetype = "dashed") +
  annotate("text", x = -8700, y = uppers + 3.5, label = as.character(uppers)) +
  annotate("text", x = -8700, y = lowers - 3.5, label = as.character(lowers)) +
  scale_x_continuous(breaks = sort(c(sdate, -6000, -3000, 0))) +
  scale_y_continuous(breaks = sort(c(0, sealevel, 100)))

@@

# Create label for map
dist_label <- paste0("Horizontal distance = ", round(hordist, 1), "m", 
                "\nTopographic distance = ", round(tdist, 1), "m",
                "\nVertical distance = ", round(vertdist, 1), "m")

@@

# Create plot showing single simulation of sea-level
seasamp <- ggplot() +
  geom_sf(data = seapoly, fill = "#dbe3f3", colour = NA) +
  geom_sf(data = sitel, fill = NA, colour = "red") +
  #labs(title =  paste0("Sea-level at ", sealevel, " m above present")) +
  coord_sf(xlim = c(st_bbox(location_bbox)[1], st_bbox(location_bbox)[3]),
           ylim = c(st_bbox(location_bbox)[2], st_bbox(location_bbox)[4]),
           expand = FALSE) +
  ggsn::scalebar(data = sitel, dist = 150, dist_unit = "m",
                 transform = FALSE, st.size = 3, height = 0.3,
                 border.size = 0.1, st.dist = 0.5,
                 anchor = c(x = anc[2] - 500, y = anc[1] + 800)) +
  # Uncomment to annotate with distance measures
  # annotate("text", x = st_bbox(location_bbox)[1] + 20,
  #          y = st_bbox(location_bbox)[4] - 100,
  #          label = dist_label, hjust = 0) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

@@

This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty. Here the approach was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and both quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

>>>>

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @åstveit2018; @solheim2020]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

```{r map-iso,  fig.cap = paste("A) Distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves. The isobases have a direction of 327$^{\\circ}$ (Romundset et al. 2018), B) Displacement curves. Note the increasing steepness of the curves towards the north-east."), fig.height = 13}
bboxsites <- st_bbox(sites_sa)
bboxsites[1] <- bboxsites[1] - 15000
bboxsites[3] <- bboxsites[3] + 15000
bboxsites[2] <- bboxsites[2] - 5000
bboxsites[4] <- bboxsites[4] + 5000
bboxsitespoly <- st_as_sf(st_as_sfc(bboxsites))

@@

anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
overview_map <- ggplot() +
  geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
  geom_sf(data = isobases, aes(colour = name)) +
  geom_sf(data = st_centroid(sites_sa), fill = "red",
          size = 2.25, shape = 21,
          colour = "black") +
  ggsn::scalebar(data = sites_sa, dist = 20, dist_unit = "km",
                 transform = FALSE, st.size = 4, height = 0.02,
                 border.size = 0.1, st.dist = 0.03,
                 anchor = c(x = anc[2] - 15500, y = anc[1]) + 8000) +
  coord_sf(xlim = c(bboxsites[1], bboxsites[3]),
           ylim = c(bboxsites[2], bboxsites[4]),
           expand = FALSE) +
  scale_colour_manual(values = c("Arendal" = "black",
                                   "Larvik" = "darkgreen",
                                   "Tvedestrand" = "blue",
                                   "Horten" = "darkorange")) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

@@

To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done for each site by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. These values were then subtracted from the displacement curve located to the south-west of the site, and in the few cases where the site was located south-west of all isobases, the values were instead added. The result of this process is shown for an example site in Figure \@ref(fig:example-1).

@@

# Plot the drawn elevation on the displacement curve
elevsamp <- ggplot(sitecurve, aes(x = years)) +
  geom_line(aes(y =  upperelev), colour = "red") +
  geom_line(aes(y = lowerelev), colour = "red") +
  ylab("Meters above present sea-level") +
  xlab("cal BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal") +
  geom_segment(aes(x = sdate, xend = sdate, y = uppers,
                                           yend = -Inf), col = "black",
                size = 0.25, linetype = "solid") +
  geom_segment(aes(x = -Inf, xend = sdate, y = uppers,
                   yend = uppers), col = "black", size = 0.25) +
  geom_segment(aes(x = -Inf, xend = sdate, y = lowers,
                   yend = lowers), col = "black", size = 0.25) +
  geom_segment(aes(x = -Inf, xend = sdate, y = lowers,
                   yend = lowers), col = "black", size = 0.25) +
  geom_segment(aes(x = -Inf, xend = sdate, y = sealevel,
                   yend = sealevel), col = "red", size = 0.25,
               linetype = "dashed") +
  annotate("text", x = -8700, y = uppers + 3.5, label = as.character(uppers)) +
  annotate("text", x = -8700, y = lowers - 3.5, label = as.character(lowers)) +
  scale_x_continuous(breaks = sort(c(sdate, -6000, -3000, 0))) +
  scale_y_continuous(breaks = sort(c(0, sealevel, 100)))

@@

# Create hillshade
# For some reason the terra version returned pathchy hillshade (probably due 
# to the defaults - I didn't bother experimenting too much as this is simply for
# the purposes of visualisation).
slope <- raster::terrain(raster(sitearea), 'slope')
aspect <- raster::terrain(raster(sitearea), 'aspect')
hill <- raster::hillShade(slope, aspect, 40, 270)

# Make the raster amenable for plotting with ggplot (as.data.frame with terra
# returned some strange errors)
raster_df <- raster::as.data.frame(raster(sitearea), xy = TRUE)
names(raster_df) <- c("x", "y", "value")

@@

# Create plot showing single simulation of sea-level
seasamp <- ggplot() +
  geom_raster(data = raster::as.data.frame(hill, xy = TRUE),
                aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "black", high = "grey40", na.value = NA) +
  new_scale_fill() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = value),
                alpha = 0.3) +
  scale_fill_gradient(low = "grey", high = "white", na.value = NA) +
  new_scale_fill() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
  scale_fill_gradient(low = NA, high = NA, na.value = "white") +
  geom_sf(data = seapoly, fill = "#dbe3f3", colour = NA) +
  geom_sf(data = sitel, fill = NA, colour = "red") +
  #labs(title =  paste0("Sea-level at ", sealevel, " m above present")) +
  coord_sf(xlim = c(st_bbox(location_bbox)[1], st_bbox(location_bbox)[3]),
           ylim = c(st_bbox(location_bbox)[2], st_bbox(location_bbox)[4]),
           expand = FALSE) +
  ggsn::scalebar(data = sitel, dist = 150, dist_unit = "m",
                 transform = FALSE, st.size = 3, height = 0.3,
                 border.size = 0.1, st.dist = 0.5,
                 anchor = c(x = anc[2] - 500, y = anc[1] + 800)) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

# Create label of distance measures
dist_label <- paste0("Horizontal distance = ", round(hordist, 1), "m", 
                "\nTopographic distance = ", round(tdist, 1), "m",
                "\nVertical distance = ", round(vertdist, 1), "m")

@@

This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty. Here the approach was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
